Divergent elliptical biface morphology: Distinct local retouch practice or serpent dance proxy for Elapids and Vipers?

### 2022-12-20 supp/index.qmd
<edit index>

Elliptical bifaces are prevalent at Mississipian sites throughout the American Southeast. Those from Millsap Cache and Jowell Farm comprise two of the largest samples of this ill-understood stone tool from the ancestral Caddo area. Bifaces from Millsap Cache were reportedly produced using Kay County flint, while those from Jowell Farm were manufactured using Edwards chert, providing for an empirical test of morphological differences as a function of raw material. The sample of elliptical bifaces was divided into two size classes; one conceptually reflective of production (large), and the other with local reduction practices (small). Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface shapes in the small class. Size classes were also used to test the hypothesis that greater morphological variation would be apparent in the small class due to idiosyncratic responses related to local retouch practices. Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting that a shared and morphologically-consistent mental template was maintained independent of biface size. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. This finding supports the argument that elliptical biface morphology diverges through local reduction practices. As expected, greater shape diversity occurs in the small class, where Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Distinct local reduction practices are advanced as the driver of extant morphological differences found in elliptical bifaces.

@@

@@

Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture from Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance. It is plausible that in this case biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species. Differences in biface **size** may articulate with the size of a specific species, as some are smaller than others (i.e. _Sistrurus miliarius streckeri_ or western pygmy rattlesnake). Alternatively, **size** differences might also depict allometric trends associated with the development and maturation of distinct species, as well as sexual size dimorphism.

```{r figspiro, fig.cap="Shell cup from Spiro Mound depicting the serpent dance, where the difference in elliptical biface morphology between snake dancers and eagle dancers is made plain. This figure was redrafted from Figure 44 in the Phillips and Brown series [-@RN2447]."}

@@

This study uses the tools of landmark geometric morphometrics to assess whether resharpening and/or retouch of these important ceremonial bifaces may have resulted in convergent morphology at two sites in the ancestral Caddo area. A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more diverse at the larger size, suggestive of greater diversity in biface shape or size related to production activities.

Results can also be used to evaluate the hypothesis that elliptical biface morphology is representative of local species of the Elapidae and Viperidae families, where greater diversity would be expected in the larger size class. If representative of local snake species, the analysis of morphological disparity would indicate a greater diversity of size in the larger size class. This is due to the fact that all species will traverse the smaller size class during maturation and development before reaching adulthood, where morphology is known to differ across species.

## Analysis overview

@@

### Generalised Procrustes Analysis

@@

### Principal Components Analysis

@@

### Procrustes ANOVA

@@

## Primary findings

  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material.
  -   Allometry (shape change with size) was not significant.
  -   Bifaces in the _large class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large class_, supporting arguments for production-based standardization.
  -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch or retouch intensity.
  -   Morphological disparity in the shape of the _small class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache. This finding provides additional support for the hypothesis that greater shape diversity is associated with retouch efforts, rather than production.

## Parallel-oblique flaking

@@

## Summary

@@

## Acknowledgments {-}

@@

## Funding {-}

@@

## Data management {-}

@@

## Colophon {-}

>>>>

## Background

Elliptical bifaces are prevalent at Mississipian and Caddo sites throughout the American Southeast, and are depicted in motifs that adorn shell gorgets and bowls, representing two distinct dances. Morphological attributes of elliptical bifaces used in Mississipian and Caddo imagery differs in noticeable ways that are manifest in associated material culture elements, and bifaces from Caddo sites more readily articulate with the serpent dance, where the motif portrays the emergence of the Hero Twins from the Beneath World; each grasping an elliptical biface.

## Hypotheses

The elliptical bifaces used in this study represent two of the largest samples cached at Caddo sites, and each was manufactured with a distinct raw material, providing for an empirical test of morphological differences as a function of raw material. If the elliptical bifaces were intended to represent local elapid and viper populations in the serpent dance, then their morphology may differ due to regional differences in biogeography. The sample of elliptical bifaces was also divided into two size classes; one conceptually reflective of production (large) or adult elapids and vipers, and the other with local reduction practices (small) or juvenile elapids and vipers.

Size classes were used to assess whether modifications by Caddo knappers may have yielded similar---convergent---biface morphologies in the small class suggestive of relative standardization due to resharpening and/or retouch in one instance, and morphological divergence associated with adult elapids and vipers in the other. Size classes were also used to test two hypotheses related to morphological diversity; one where greater morphological variation would be apparent in the small size class due to idiosyncratic responses related to local retouch practices, and a second where greater morphological variation would be apparent in the large size class due to differences in constituent elapid and viper communities depicted by the bifaces.

## Results

Results demonstrate that elliptical biface shape does not differ by raw material, but size does, suggesting that a shared---and morphologically-consistent---mental template was maintained independent of biface size and constituent community differences in elapids and vipers. The subsequent analysis of elliptical biface morphology by size class demonstrated that size does not differ by raw material in the large class, but in the small class, it does. These findings support the argument that elliptical biface morphology diverges through local reduction practices, while simultaneously nullifying the hypothesis that biface shape reflects the morphology of local elapid and viper communities. As expected, greater shape diversity occurs in the small size class, where the Jowell Farm bifaces were found to be more morphologically diverse than those from Millsap Cache. Thus, distinct local reduction practices are advanced as the driver of extant morphological differences expressed in elliptical bifaces.

@@

# Background

@@

Among the shell cup illustrations assembled by Phillips and Brown [-@RN2447] that characterize the material culture at Spiro Mound is a depiction of elliptical bifaces conveyed as elements of the serpent/snake dance rather than elements of the eagle/bird dance. It is plausible that in this case biface **shape** may convey a relevant---and ceremonially important---representation related to extant morphological differences expressed in local elapid and viper species. Differences in biface **size** may articulate with the size of a specific species, as some are smaller than others (i.e. _Sistrurus miliarius streckeri_ or western pygmy rattlesnake). Alternatively, **size** differences might also depict allometric trends associated with the development and maturation of distinct species, as well as sexual size dimorphism.

```{r figspiro, fig.cap="Shell cup from Spiro Mound depicting the serpent dance, where the difference in elliptical biface morphology between eagle dancers and snake dancers is made plain. This figure was redrafted from Figure 44 in the Phillips and Brown series [-@RN2447]."}

@@

## Hypotheses

1.   Elliptical bifaces found at Millsap Cache were manufactured using Kay County Chert, while those from Jowell Farm were manufactured using Edwards Chert. Differences in stone tool shape as a function of raw material have been reported for Perdiz arrow points in the region [@RN9364]; however, others have demonstrated that the attribution of morphological variability to raw material differences is an assumption that warrants regular testing, suggesting that primacy should not be assumed [@RN8442]. Thus, a comparison of between-site morphology was included to assess elliptical biface morphology as a function of raw material.
  a)    The biogeography of elapid and viper species differs within the ancestral Caddo area, where elapid populations occur south of the Red River, excluding them from the geography surrounding Millsap Cache. If elliptical biface morphology is intended to represent local elapid and viper populations, then elliptical bifaces from these sites will differ due to differences in the constituent communities of elapids and vipers.

2.    The elliptical bifaces from Jowell Farm were discovered in an area of the ancestral Caddo region where others have assumed large bifaces to be products of external manufacture due to the absence of local production evidence and the availability of raw material/s of similar size and quality [@RN3682;@RN11783;@RN21001]. Given the potential that elliptical bifaces were produced by knappers outside of the ancestral Caddo region for at least one of the caches (Jowell Farm), it is assumed that the smaller elliptical bifaces will more readily reflect morphology imposed by Caddo knappers through resharpening and/or retouch incurred through normal use [@RN11264]. Thus, it is expected that elliptical bifaces will express convergent morphology, where morphological consistency is achieved through the idiosyncratic approaches to resharpening and/or retouch by Caddo knappers in the small size class.
  a)   Elapid and viper species will traverse the smaller size class through normal developmental stages before reaching adulthood, where their morphology is known to differ within and among species. If elliptical biface morphology is thought to represent distinct elapid and viper populations, then the most pronounced morphological difference will occur in the large size class among the adults. Thus, if elliptical bifaces were intended to represent elapid and viper species, then the morphological pattern will be identical to H2. However, the analytical results would be read in reverse, where expected similarities occur at the smaller size class, before diverging in the large size class.

3.    A measure of morphological disparity was subsequently used to assess whether biface morphology may have been more morphologically diverse in the large size class, suggestive of greater diversity in biface shape or size related to production activities [@RN11264]. An alternative hypothesis is that the idiosyncratic responses to resharpening and/or retouch by Caddo knappers required to address damage incurred through normal use generate greater morphological diversity in the small size class [@RN11264].
  a)    Elapid and viper species express greater morphological diversity at their larger size. Thus, if elliptical bifaces were intended to represent elapid and viper species, and biological diversity differs between sites---elapid biogeography restricted to the area of Jowell Farm---then morphological disparity will occur between sites in the large size class.

# Analysis overview

@@

## Generalised Procrustes Analysis

@@

## Principal Components Analysis

@@

## Procrustes ANOVA

@@

# Primary findings

  -   Allometry (shape change with size) was not significant.
  -   Bifaces _do not differ_ in shape, but **differ significantly** in size by raw material.
  -   Bifaces in the _large class_ recovered from Millsap Cache and Jowell Farm do not differ in shape or size, suggesting consistent morphology in bifaces from the _large class_, supporting arguments for relative production-based standardization.
  -   Bifaces in the _small class_ recovered from Millsap Cache and Jowell Farm **differ significantly** in size. Thus, while shape remains consistent, size differs, suggesting---at minimum---locally divergent approaches to retouch or retouch _intensity_.
  -   Morphological disparity in the **shape** of elliptical bifaces in the _small size class_ is **significantly greater** at Jowell Farm, providing evidence for a more standardized approach to local retouch at Millsap Cache. This finding also yields added support for the hypothesis that greater shape diversity is associated with local retouch efforts, rather than production.

# Parallel-oblique flaking

@@

In the case of ceremonial bifaces, the intentional use of parallel oblique flake scars may have been intended to convey motion, where flake scars on one face represent muscle constriction on one edge and muscle dilation on the other, with an inverted pattern on the opposite face. Thus, in intentionally flipping between the two faces---which might have been a component of the serpent/snake dance---dancers could convey motion with a static implement. Additionally, the parallel oblique flake scars closely mimic the parallel scales that occur beneath both elapids and vipers, which shift in similar ways to generate motion.

# Summary

@@

# Acknowledgments {-}

@@

# Funding {-}

@@

# Data management {-}

@@

# Colophon {-}
